---
title: "Storm Data Analysis"
author: "Humbert Costas"
date: "Sunday, January 25, 2015"
output: html_document
---
# Synopsis
Storms and other severe weather events can cause both public health and economic problems for communities and municipalities. Many severe events can result in fatalities, injuries, and property damage, and preventing such outcomes to the extent possible is a key concern.

This project involves exploring the U.S. National Oceanic and Atmospheric Administration's ([NOAA](http://www.noaa.gov/)) storm [database](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2FNCDC%20Storm%20Events-FAQ%20Page.pdf). This database tracks characteristics of major storms and weather events in the United States, including when and where they occur, as well as estimates of any fatalities, injuries, and property damage.

# Data Processing
You can download the [storm data](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2) used in this analysis, to check what's going on in this paper.

This code puts the data in *rawdata* variable:
```{r cache=TRUE}
# download.file(url='https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2', destfile='stormdata.csv.bz2')
rawdata <- read.csv('stormdata.csv.bz2')
```

Vamos a analizar los datos empezando por sus dimensiones. El conjunto de datos comprende `r nrow(rawdata)` observaciones de `r ncol(rawdata)` variables cada una, concretamente:

```{r}
names(rawdata)
```

En este analisis nos vamos a fijar en los tipos de eventos (EVTYPE) registrados y las variables relacionadas con la salud de la población y costes económicos.

La clasificación de los eventos metereologicos no parece seguir una clasificación muy cuidada. Se trata de `r length(levels(rawdata$EVTYPE))` tipos distintos de eventos, por ejemplo, veamos los 20 ultimos: 

```{r echo=FALSE}
events <- levels(rawdata$EVTYPE)
tail(events,20)
```

Antes de trabajar con estos datos vamos a convertir los *EVTYPE* a mayusculas. 

```{r}
events <- as.data.frame(toupper(events))
names(events)[1] = "EVTYPE"
```

Una vez normalizados los textos, se categorizan agregarando los eventos más genericos, tales como HURRICANE, HEAT, WIND, FLOOD, etc. 

```{r echo=FALSE, cache=TRUE}
# Para categorizar eventos usaremos la variable category
category_filter <- ".*TSTM.*|.*THUNDERSTORM.*"
category_name <- "THUNDERSTORM WIND"
category <- as.character(events[1:nrow(events) %in% grep(category_filter, events$EVTYPE),])
rawdata$EVTYPE[rawdata$EVTYPE %in% category] <- category_name

category_filter <- ".*URRICAN.*|.*TYPHOON.*"
category_name <- "HURRICANE"
category <- as.character(events[1:nrow(events) %in% grep(category_filter, events$EVTYPE),])
rawdata$EVTYPE[rawdata$EVTYPE %in% category] <- category_name

category_filter <- ".*WILD.*FIRE.*"
category_name <- "WILDFIRE"
category <- as.character(events[1:nrow(events) %in% grep(category_filter, events$EVTYPE),])
rawdata$EVTYPE[rawdata$EVTYPE %in% category] <- category_name

category_filter <- ".*EX.*HEAT.*"
category_name <- "HEAT"
category <- as.character(events[1:nrow(events) %in% grep(category_filter, events$EVTYPE),])
rawdata$EVTYPE[rawdata$EVTYPE %in% category] <- category_name

category_filter <- ".*WINTER.*STORM.*"
category_name <- "WINTER STORM"
category <- as.character(events[1:nrow(events) %in% grep(category_filter, events$EVTYPE),])
rawdata$EVTYPE[rawdata$EVTYPE %in% category] <- category_name

category_filter <- ".*COLD.*|.*FROST.*|.*FREEZE.*"
category_name <- "COLD"
category <- as.character(events[1:nrow(events) %in% grep(category_filter, events$EVTYPE),])
rawdata$EVTYPE[rawdata$EVTYPE %in% category] <- category_name

category_filter <- "(HIGH|STRONG|BLOWING).*WIND.*"
category_name <- "WIND"
category <- as.character(events[1:nrow(events) %in% grep(category_filter, events$EVTYPE),])
rawdata$EVTYPE[rawdata$EVTYPE %in% category] <- category_name

category_filter <- ".*FLOOD.*"
category_name <- "FLOOD"
category <- as.character(events[1:nrow(events) %in% grep(category_filter, events$EVTYPE),])
rawdata$EVTYPE[rawdata$EVTYPE %in% category] <- category_name

category_filter <- ".*HEAVY.*RAIN.*"
category_name <- "HEAVY RAIN"
category <- as.character(events[1:nrow(events) %in% grep(category_filter, events$EVTYPE),])
rawdata$EVTYPE[rawdata$EVTYPE %in% category] <- category_name

category_filter <- ".*TROPICAL.*STORM.*"
category_name <- "TROPICAL STORM"
category <- as.character(events[1:nrow(events) %in% grep(category_filter, events$EVTYPE),])
rawdata$EVTYPE[rawdata$EVTYPE %in% category] <- category_name

category_filter <- ".*RIP CURRENT.*"
category_name <- "RIP CURRENTS"
category <- as.character(events[1:nrow(events) %in% grep(category_filter, events$EVTYPE),])
rawdata$EVTYPE[rawdata$EVTYPE %in% category] <- category_name

category_filter <- ".*AVALANCH.*"
category_name <- "AVALANCHE"
category <- as.character(events[1:nrow(events) %in% grep(category_filter, events$EVTYPE),])
rawdata$EVTYPE[rawdata$EVTYPE %in% category] <- category_name
```

# Questions:

## Across the United States, which types of events are most harmful with respect to population health?
```{r cache=TRUE}
library(ggplot2)
human_damages <- aggregate(cbind(FATALITIES, INJURIES) ~ EVTYPE, rawdata, sum)
human_damages$TOTAL <- human_damages$FATALITIES + human_damages$INJURIES
quest1 <- head(human_damages[order(-human_damages$TOTAL, -human_damages$FATALITIES, -human_damages$INJURIES), ], 10)
qplot(x=TOTAL, y=EVTYPE, data = quest1)
```

```{r}
#rawdata$PROPDMG
```
# Results